---
title: "Principal Component Analysis using R"
output:
  md_document:
    variant: markdown_github
---

The current markdown describes Principal Component Analysis (PCA) on the dataset: Measurement of bones and ecological group of birds, available in [Kaggle](https://www.kaggle.com/zhangjuefei/birds-bones-and-living-habits)

In total, there are 6 ecological groups and 10 measurements to represent each bird.

### Ecological Group
* Swimming Birds
* Wading Birds
* Terrestrial Birds
* Raptors
* Scansorial Birds
* Singing Birds


### Measurement variables
* Length and Diameter of Humerus (huml, humw)
* Length and Diameter of Ulna (ulnal, ulnaw)
* Length and Diameter of Femur (feml, femw)
* Length and Diameter of Tibiotarsus (tibl, tibw)
* Length and Diameter of Tarsometatarsus (tarl, tarw)
        

```{r}
library(dlookr)
library(ggplot2)
library(psych)
library(imputeTS)
```


```{r}
birds <- read.csv("bird.csv", header=TRUE)
head(birds)
summary(birds)
birds$type<-as.factor(birds$type)
```
### Addressing missing values 

If you get the below error, it is likely that you have missing values in your data. 

> Error in svd(x, nu = 0, nv = k) : infinite or missing values in 'x'

A pareto chat is created to visualize the missing values and address them.

```{r}
plot_na_pareto(birds)
```

You could use multiple ways to address the missing values. Machine learning methods such as MICE, rpart could be used. In this scenario, I am addressing them simply by removing them which is not recommended at all times. I will address this aspect with other methods and draw comparison in future iterations.
```{r}
#Handling of missing values, substitution with mean
birds <- na.omit(birds)
dims.birds <- birds[,2:11]
#dims.birds <- na_mean(dims.birds)
```

### Tests to determine appropriateness of the dataset for PCA

```{r}
KMO(dims.birds)
```

The KMO seems to be above 0.80 (meritorius : Kaiser, H. F., & Rice, J. (1974). Little jiffy, mark IV. Educational and psychological measurement, 34(1), 111-117. ) for all values denoting the data is great for PCA and factor analysis. 

```{r}
bartlett.test(dims.birds) #multiple variances
```
The Bartlett test of homegenity of variances for multiple variances shows a significant result (p<0.05), indicating appropriateness of the dataset for PCA; rejection of null hypothesis - correlation matrix is an identity matrix. 

### Plotting a scree plot

Showcases the graph between eigen values and components
```{r}
pca.birds <- prcomp(dims.birds, scale=TRUE)
plot(pca.birds, type="l")

```


### Summary values
```{r}
summary.birds<-summary(pca.birds)
summary.birds

```

### Biplot

Biplot showcases the PC scores with the loadings of variables

```{r}
final.birds<- cbind(birds, pca.birds$x[,1:2])
biplot(pca.birds, scale=0)
```

### Plotting the confidence ellipses and variance explained

```{r}
final.birds %>%
  ggplot(aes(PC1, PC2, col=type, fill=type))+
  stat_ellipse(geom = "polygon", alpha=0.5)+
  geom_point()+
  xlab(paste('PC1 - ',round(summary.birds$importance[2,1],2)*100, '% variance explained'))+
  ylab(paste('PC2 - ',round(summary.birds$importance[2,2],2)*100,'% variance explained'))
```


